Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2021-07-12 edubots_surveys.Rnw
Fixes during conference

\begin{abstract} % abstract
Introducing new technologies such as messaging platforms, and the chatbots attached to them, in higher education, is rapidly growing. This introduction entails a careful consideration of the potential opportunities and/or challenges of adopting these tools. Hence, a thorough examination of the teachers’ experiences in this discipline can shed the light on the effective ways of enhancing students’ learning and boosting their progress. In this contribution, we have surveyed the opinions of tertiary education teachers based mainly in Spain and Spanish-speaking countries. The focus of these surveys is to collect teachers’ feedback about their opinions of the introduction of the messaging platforms and chatbots in their classes, understand their needs and to gather information about the various educational use cases where these tools are valuable. In addition, an analysis of how and when teachers’ attitudes towards the use of these tools can vary across gender, experience, and their discipline of specialisation is presented. The key findings of this study highlight the factors that can contribute to the advancement of the adoption of messaging platforms and chatbots in higher education institutions to achieve the desired learning outcomes.
\end{abstract}

@@

% TODO: introduction
The introduction of new technologies in the classroom involves a combination of teacher training, student enthusiasm and possibly acquiring new skills by both. Some technologies are readily adapted, but others require longer time for full adoption. In most cases, the magnitude of the uptake of the new technology impinges on the collaboration of all the parties using it. Hence, examining the users’ attitudes and their opinions is an important first step in boosting this collaboration and reaping the benefits of the technology.  Recently, a surge in the use of synchronous instant messaging tools in higher education such as WhatsApp, Telegram and Facebook Messenger took place. Some messaging applications have built in chatbots that are able to support a synchronous conversation between the different parties \cite{differ.chat}, while other chatbots are developed as standalone systems and can possibly be attached to a messaging application using an API (Application Program Interface).

@@

WhatsApp is, at least in most Western countries,
the most popular MIM platform used by educators to give assignments'
feedback to students, support course discussions, and provide learning
resources in informal learning settings \cite{panah2020study}.  Moreover, the
use of WhatsApp in higher education could enhance social presence
\cite{tang2017mobile} and foster trust relationships between educators
and students embedded in the social learning process
\cite{gachago2015crossing}; however, this last paper also reflects the need for
learners to "take ownership of the tool" and the advantages of social learning
in general. At the same time, it also mentions different challenges, among which
the most important is the blurring of social and academic life. Examining these
attitudes will be the objective of RQ3 in this paper.

Nevertheless, there are challenges when using MIM tools that occur due
to the blurring of boundaries between academic and private life. This
can lead to technostress \cite{gachago2015crossing}, difficulty in
managing responsibilities, especially among mature students, and lack of
privacy \cite{tang2017mobile}.  Students' dropout of the MIM groups,
as they can leave groups at any time, can hinder their learning and
undermine educators' efforts \cite{mwakapina2016whatsapp}.  In
addition, there is a need to set rules and norms for these MIM groups
in order to maintain the safety of these online communities for
students \cite{abbas2021onlinechat}. However, these rules should not affect students'
ownership and control, since it is vital to advance in their learning
\cite{gachago2015crossing} process. 
This is why examining the role of MIM in higher education is still a challenge,
and why the attitudes of the teaching community towards them have to be
examined, as we do in this paper.

@@

Along this line of research, our previous work \cite{MoraChatbots2021} aimed at analysing the expectations of students on this regard; recent papers have also analyzed how the use of chatbots affected the learning outcomes of students in a Chinese class \cite{doi:10.1177/0735633120929622}. This paper analyzed conversational chatbots in an one-to-one setting, finding that learning really benefited from it. This is, once again, a proof of the benefit of chatbots in certain settings; however, there are some attitudes that might prevent the successful acceptance of the technology, and these are what we are trying to find out in this paper, together with what kind of features would make its acceptance better.

@@

We will present next the methodology we have followed to find out teachers' attitudes.

@@

\section{Results}
\label{sec:res}

@@

For answering RQ1, teachers were queried about whether they use messaging apps in their classes, specifically, Telegram, WhatsApp, Slack, or any other messaging app, and whether they use messaging apps provided by their Academic Institution. 
According to the results summarized in Fig. \ref{figure:chatbots_use}, teachers mainly use the instant messaging app provided by their institutions (55\%),  and Whatsapp (44\%), the most popular messaging app in Spain. Finally, about 20\% also use Telegram. Only 19 teachers (6.7%) replied that they do not use any messaging app in their class. 
Although gender differences are not significant, female teachers answered they use instant messaging apps more than male teachers (about 10\% more). Also, teachers in vocational Education also use Whatsapp more than university teachers.

@@

Some of the questions in the survey were focused on the impact of the COVID19 pandemic between 2020 and 2021, in the teachers' attitudes towards the instant messaging apps. The 273 answers are summarized in Fig. \ref{figure:chatbots_use_postcovid}, showing that about 77\% of teachers already used these tools before the pandemic and kept using them during the pandemic lockdowns that forced students and educators to use remote education schemes. Moreover, approximately 15\% of them switched their messaging app for one that offered a safer interaction with their students. According to the responses, an additional 16\% started using messaging apps during the pandemic for the first time in their classes. Aggregating the answers, currently 91\% of the teachers are using messaging apps as learning tools.

@@

The key findings of this study shed the light on educators’ preference to use messaging applications that their institutions support.  Technology adoption literature often focus on users’ perceptions of the technology’s usefulness and ease of use as important prerequisites for successful adoption and utilization. Nevertheless, in higher education, institutions’ role in integrating these tools to their educational systems can improve the uptake of these applications and shape the social and educational experiences of their students.  This integration, to be fruitful, institutions have to ensure that these messaging applications not only are GDPR compliant to keep students’ data secure but also should provide IT support to all stakeholders who use the applications .  
Moreover, this study clarifies the differences between the teachers’ perspectives and/or intentions and that of their students when using the messaging applications in higher education.  Teachers tend to adopt technologies supported by their institutions.  This could be to ensure that their universities can oversee their efforts in supporting their students during the learning process.  Another reason could be their familiarity with the technology provided by their institutions. On the other hand, students use the non-institution messaging applications to form informal discussion groups with their peers.  It is worth noting that peer support and collaboration is inseparable from learning \cite{timmis2012constant} and correlates positively with higher retention in higher education \cite{o2014front}.  Therefore, both perspectives are complementary and play different roles in promoting the learning process.
%This section is repeated
%\section*{Acknowledgements}

@@

% Antonio - If there is any, please indicate.

>>>>

\begin{abstract} % abstract
Introducing new technologies such as messaging platforms, and the chatbots attached to them, in higher education, is rapidly growing. This introduction entails a careful consideration of the potential opportunities and/or challenges of adopting these tools. Hence, a thorough examination of the teachers’ experiences in this discipline can shed the light on the effective ways of enhancing students’ learning and boosting their progress. In this contribution, we have surveyed the opinions of tertiary education teachers based mainly in Spain and Spanish-speaking countries. The focus of these surveys is to collect teachers’ feedback about their opinions of the introduction of the messaging platforms and chatbots in their classes, understand their needs and to gather information about the various educational use cases where these tools are valuable. In addition, an analysis of how and when teachers’ opinions towards the use of these tools can vary across gender, experience, and their discipline of specialisation is presented. The key findings of this study highlight the factors that can contribute to the advancement of the adoption of messaging platforms and chatbots in higher education institutions to achieve the desired learning outcomes.
\end{abstract}

@@

% TODO: introduction
The introduction of new technologies in the classroom involves a combination of teacher training, student enthusiasm and possibly acquiring new skills by both. Some technologies are readily adapted, but others require longer time for full adoption. In most cases, the magnitude of the uptake of the new technology impinges on the collaboration of all the parties using it. Hence, examining the users’ opinions and their opinions is an important first step in boosting this collaboration and reaping the benefits of the technology.  Recently, a surge in the use of synchronous instant messaging tools in higher education such as WhatsApp, Telegram and Facebook Messenger took place. Some messaging applications have built in chatbots that are able to support a synchronous conversation between the different parties \cite{differ.chat}, while other chatbots are developed as standalone systems and can possibly be attached to a messaging application using an API (Application Program Interface).

@@

WhatsApp is, at least in most Western countries,
the most popular MIM platform used by educators to give assignments'
feedback to students, support course discussions, and provide learning
resources in informal learning settings \cite{panah2020study}.  Moreover, the
use of WhatsApp in higher education could enhance social presence
\cite{tang2017mobile} and foster trust relationships between educators
and students embedded in the social learning process
\cite{gachago2015crossing}; however, this last paper also reflects the need for
learners to "take ownership of the tool" and the advantages of social learning
in general. At the same time, it also mentions different challenges, among which
the most important is the blurring of social and academic life. Examining these
opinions will be the objective of RQ3 in this paper.

Nevertheless, there are challenges when using MIM tools that occur due
to the blurring of boundaries between academic and private life. This
can lead to technostress \cite{gachago2015crossing}, difficulty in
managing responsibilities, especially among mature students, and lack of
privacy \cite{tang2017mobile}.  Students' dropout of the MIM groups,
as they can leave groups at any time, can hinder their learning and
undermine educators' efforts \cite{mwakapina2016whatsapp}.  In
addition, there is a need to set rules and norms for these MIM groups
in order to maintain the safety of these online communities for
students \cite{abbas2021onlinechat}. However, these rules should not affect students'
ownership and control, since it is vital to advance in their learning
\cite{gachago2015crossing} process. 
This is why examining the role of MIM in higher education is still a challenge,
and why the opinions of the teaching community towards them have to be
examined, as we do in this paper.

@@

Along this line of research, our previous work \cite{MoraChatbots2021} aimed at analysing the expectations of students on this regard; recent papers have also analyzed how the use of chatbots affected the learning outcomes of students in a Chinese class \cite{doi:10.1177/0735633120929622}. This paper analyzed conversational chatbots in an one-to-one setting, finding that learning really benefited from it. This is, once again, a proof of the benefit of chatbots in certain settings; however, there are some opinions that might prevent the successful acceptance of the technology, and these are what we are trying to find out in this paper, together with what kind of features would make its acceptance better.

@@

We will present next the methodology we have followed to find out teachers' opinions.

@@

\section{Results and analysis}
\label{sec:res}

@@

For answering RQ1, teachers were queried about whether they use messaging apps in their classes, specifically, Telegram, WhatsApp, Slack, or any other messaging app, and whether they use messaging apps provided by their Academic Institution. 
According to the results summarized in Fig. \ref{figure:chatbots_use}, teachers mainly use the instant messaging app provided by their institutions (55\%),  and Whatsapp (44\%), the most popular messaging app in Spain. Finally, about 20\% also use Telegram. Only 19 teachers (6.7\%) replied that they do not use any messaging app in their class. 
Although gender differences are not significant, female teachers answered they use instant messaging apps more than male teachers (about 10\% more). Also, teachers in vocational Education also use Whatsapp more than university teachers.

@@

Some of the questions in the survey were focused on the impact of the COVID19 pandemic between 2020 and 2021, in the teachers' opinions towards the instant messaging apps. The 273 answers are summarized in Fig. \ref{figure:chatbots_use_postcovid}, showing that about 77\% of teachers already used these tools before the pandemic and kept using them during the pandemic lockdowns that forced students and educators to use remote education schemes. Moreover, approximately 15\% of them switched their messaging app for one that offered a safer interaction with their students. According to the responses, an additional 16\% started using messaging apps during the pandemic for the first time in their classes. Aggregating the answers, currently 91\% of the teachers are using messaging apps as learning tools.

@@

The key findings of this study shed the light on educators’ preference to use messaging applications that their institutions support.  Technology adoption literature often focus on users’ perceptions of the technology’s usefulness and ease of use as important prerequisites for successful adoption and utilization. Nevertheless, in higher education, institutions’ role in integrating these tools to their educational systems can improve the uptake of these applications and shape the social and educational experiences of their students.  This integration, to be fruitful, institutions have to ensure that these messaging applications not only are GDPR compliant to keep students’ data secure but also should provide IT support to all stakeholders who use the applications.

Moreover, this study clarifies the differences between the teachers’ perspectives and/or intentions and that of their students when using the messaging applications in higher education.  Teachers tend to adopt technologies supported by their institutions.  This could be to ensure that their universities can oversee their efforts in supporting their students during the learning process.  Another reason could be their familiarity with the technology provided by their institutions. On the other hand, students use the non-institution messaging applications to form informal discussion groups with their peers.  It is worth noting that peer support and collaboration is inseparable from learning \cite{timmis2012constant} and correlates positively with higher retention in higher education \cite{o2014front}.  Therefore, both perspectives are complementary and play different roles in promoting the learning process.

% What kidn of chatbots are preferred? (suggested by Fran)

% No remarkable differences between different groups.

% Only some COVID-response relationship.

% Future study (suggested by Noorhan): how COVID influenced adoption of these technologies.

% Emotial, mood for future work... Chatbots perform sentiment analysis

%This section is repeated
%\section*{Acknowledgements}

@@
